---
title: "Mouse Neurovascular scRNAseq: DE Gene Pathway Enrichment Analysis"
author: "Joshua Samuels"
date: "1/4/2022"
output: html_document
---

Pathway enrichment analysis for DE genes in each cluster performed with Reactome and GO: Biological Process. Analysis data downloaded as CSV file and loaded into R.
Reactome settings: Project to Human = YES, Include Interactors = NO.


### Load Libraries

```{r message=FALSE, warning=FALSE}
# Load libraries
library(SingleCellExperiment)
library(Seurat)
library(tidyverse)
library(Matrix)
library(scales)
library(cowplot)
library(RCurl)
library(BiocManager)
library(ggplot2)
library(dplyr)
library(magrittr)
library(patchwork)
library(ggraph)
library(clustree)
library(reshape2)
```


### Pathway Analysis with a Matrix View
# The full lists of significant (FDR < 0.05) Reactome and GO: BP pathways were focused to key pathways of interest. To do this, create a .csv file for each cell type of interest (one for UPregulated pathways and one for DOWNregulated pathways) with the follwing headers: Term_ID, Term_Name, and FDR. Save these as .csv files and use the following code to read the lists into R, manupulate them, and generate ordered pathway analysis heatmaps.

```{r}
# Up Regulated Data Lists
fp1 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/Cap_EC_Up.csv"
fp2 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/Arterial_EC_Up.csv"
fp3 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/ChPlex_Up.csv"
fp4 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/HomeoMG_Up.csv"
fp5 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/MG_Up.csv"
fp6 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/Venous_EC_Up.csv"
fp7 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/aSMC_Up.csv"


# Down Regulated Data Lists
fp8 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/Arterial_EC_Down.csv"
fp9 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/ChPlex_Down.csv"
fp10 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/HomeoMG_Down.csv"
fp11 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/MG_Down.csv"
fp12 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/Fibro_Down.csv"
# fp13 <- "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Go_Reactome_Focused_Analysis/Cap_EC_Down.csv"


CapECUp <- read.csv(file = fp1) # read the lsit into R and create the object 
CapECUp <- CapECUp %>%
  unite("Term", Term_ID:Term_Name, sep = "_") # combine the Term_ID and Term_Name columns
rownames(CapECUp) <- CapECUp[,1] # make the 1st column the rownames of the object
CapECUp <- subset(CapECUp, select = FDR) # keep only the FDR column
CapECUp <- rename(CapECUp,
                  Capillary_EC = FDR) # rename the FDR column the name of the cell type
CapECUp <- t(CapECUp) # transpose the data object
CapECUp <- as.data.frame(CapECUp) # convert the object to a data frame
CapECUp <- tibble::rownames_to_column(CapECUp, "CellType") # make the rowname into the 1st column with "CellType" as the column name

aECUP <- read.csv(file = fp2)
aECUP <- aECUP %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(aECUP) <- aECUP[,1]
aECUP <- subset(aECUP, select = FDR)
aECUP <- rename(aECUP,
                  Arterial_EC = FDR)
aECUP <- t(aECUP)
aECUP <- as.data.frame(aECUP)
aECUP <- tibble::rownames_to_column(aECUP, "CellType")

ChPlexUP <- read.csv(file = fp3)
ChPlexUP <- ChPlexUP %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(ChPlexUP) <- ChPlexUP[,1]
ChPlexUP <- subset(ChPlexUP, select = FDR)
ChPlexUP <- rename(ChPlexUP,
                  Choroid_Plexus = FDR)
ChPlexUP <- t(ChPlexUP)
ChPlexUP <- as.data.frame(ChPlexUP)
ChPlexUP <- tibble::rownames_to_column(ChPlexUP, "CellType")

HomeoMGUP <- read.csv(file = fp4)
HomeoMGUP <- HomeoMGUP %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(HomeoMGUP) <- HomeoMGUP[,1]
HomeoMGUP <- subset(HomeoMGUP, select = FDR)
HomeoMGUP <- rename(HomeoMGUP,
                  Homeostatic_Microglia = FDR)
HomeoMGUP <- t(HomeoMGUP)
HomeoMGUP <- as.data.frame(HomeoMGUP)
HomeoMGUP <- tibble::rownames_to_column(HomeoMGUP, "CellType")

MGUp <- read.csv(file = fp5)
MGUp <- MGUp %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(MGUp) <- MGUp[,1]
MGUp <- subset(MGUp, select = FDR)
MGUp <- rename(MGUp,
                  Microglia = FDR)
MGUp <- t(MGUp)
MGUp <- as.data.frame(MGUp)
MGUp <- tibble::rownames_to_column(MGUp, "CellType")

vECUp <- read.csv(file = fp6)
vECUp <- vECUp %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(vECUp) <- vECUp[,1]
vECUp <- subset(vECUp, select = FDR)
vECUp <- rename(vECUp,
                  Venous_EC = FDR)
vECUp <- t(vECUp)
vECUp <- as.data.frame(vECUp)
vECUp <- tibble::rownames_to_column(vECUp, "CellType")

aSMCUp <- read.csv(file = fp7)
aSMCUp <- aSMCUp %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(aSMCUp) <- aSMCUp[,1]
aSMCUp <- subset(aSMCUp, select = FDR)
aSMCUp <- rename(aSMCUp,
                  Arterial_SMC = FDR)
aSMCUp <- t(aSMCUp)
aSMCUp <- as.data.frame(aSMCUp)
aSMCUp <- tibble::rownames_to_column(aSMCUp, "CellType")


# Down
aECDown <- read.csv(file = fp8)
aECDown <- aECDown %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(aECDown) <- aECDown[,1]
aECDown <- subset(aECDown, select = FDR)
aECDown <- rename(aECDown,
                  Arterial_EC = FDR)
aECDown <- t(aECDown)
aECDown <- as.data.frame(aECDown)
aECDown <- tibble::rownames_to_column(aECDown, "CellType")

ChPlexDown <- read.csv(file = fp9)
ChPlexDown <- ChPlexDown %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(ChPlexDown) <- ChPlexDown[,1]
ChPlexDown <- subset(ChPlexDown, select = FDR)
ChPlexDown <- rename(ChPlexDown,
                  Choroid_Plexus = FDR)
ChPlexDown <- t(ChPlexDown)
ChPlexDown <- as.data.frame(ChPlexDown)
ChPlexDown <- tibble::rownames_to_column(ChPlexDown, "CellType")

HomeoMGDown <- read.csv(file = fp10)
HomeoMGDown <- HomeoMGDown %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(HomeoMGDown) <- HomeoMGDown[,1]
HomeoMGDown <- subset(HomeoMGDown, select = FDR)
HomeoMGDown <- rename(HomeoMGDown,
                  Homeostatic_Microglia = FDR)
HomeoMGDown <- t(HomeoMGDown)
HomeoMGDown <- as.data.frame(HomeoMGDown)
HomeoMGDown <- tibble::rownames_to_column(HomeoMGDown, "CellType")

MGDown <- read.csv(file = fp11)
MGDown <- MGDown %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(MGDown) <- MGDown[,1]
MGDown <- subset(MGDown, select = FDR)
MGDown <- rename(MGDown,
                  Microglia = FDR)
MGDown <- t(MGDown)
MGDown <- as.data.frame(MGDown)
MGDown <- tibble::rownames_to_column(MGDown, "CellType")

FibroDown <- read.csv(file = fp12)
FibroDown <- FibroDown %>%
  unite("Term", Term_ID:Term_Name, sep = "_")
rownames(FibroDown) <- FibroDown[,1]
FibroDown <- subset(FibroDown, select = FDR)
FibroDown <- rename(FibroDown,
                  Fibroblast = FDR)
FibroDown <- t(FibroDown)
FibroDown <- as.data.frame(FibroDown)
FibroDown <- tibble::rownames_to_column(FibroDown, "CellType")

# CapECDown <- read.csv(file = fp13)
# CapECDown <- CapECDown %>%
#   unite("Term", Term_ID:CapECDown, sep = "_")
# rownames(CapECDown) <- CapECDown[,1]
# CapECDown <- subset(CapECDown, select = FDR)
# CapECDown <- rename(CapECDown,
#                   Capilary_EC = FDR)
# CapECDown <- t(CapECDown)
# CapECDown <- as.data.frame(CapECDown)
# CapECDown <- tibble::rownames_to_column(CapECDown, "CellType")
 
CapECDown <- aECDown # needed a workaround for this cell type, for some reason the above code did not work for this group. This gives the same result.
 CapECDown[1, 2] <- 0.001078907
 CapECDown[1, 1] <- "Capillary_EC"


# Label manipulation prior to merging
# Here, we are renaming term names that are basically the same, removing terms that we are not interested in, and abbreviating long term names
aSMCUp <- rename(aSMCUp, 
                 "R-HSA-9613829_Chaperone Mediated Autophagy" = "R-HSA-9612973_Chaperone Mediated Autophagy",
                 "R-HSA-3371511_HSF1 activation" = "R-HSA-3371556_HSF1 activation"
                 )

vECUp$`R-HSA-449147: Signaling by Interleukins` <- NULL

ChPlexUP <- rename(ChPlexUP,
                   "R-HSA-3371511_HSF1 activation" = "R-HSA-3371571_HSF1 activation"
                   )

MGUp <- rename(MGUp,
               "GO:1904706_Neg reg of SMC proliferation" = "GO:1904706_Negative regulation of vascular associated smooth muscle cell proliferation")

CapECUp <- rename(CapECUp,
                  "R-HSA-3371511_HSF1 activation" = "R-HSA-3371453_HSF1 activation")

aECUP <- rename(aECUP,
                "R-HSA-3371511_HSF1 activation" = "R-HSA-3371453_HSF1 activation")
```

```{r}
# Merge data sets

# Up
PA.Up <- Reduce(function(x, y) merge(x, y, all = T), list(
  CapECUp,
  aECUP, 
  ChPlexUP, 
  HomeoMGUP, 
  MGUp, 
  vECUp, 
  aSMCUp)) # merge all lists and keep all unique values

# Further label manipulation post-merging
PA.Up <- rename(PA.Up,
                "R-HSA-3371497_HSP90 chaperone cycle for SHR" = "R-HSA-3371497_HSP90 chaperone cycle for steroid hormone receptors (SHR) in the presence of ligand"
                )

PA.Up$`GO:1903748_Negative regulation of establishment of protein localization to mitochondrion` <- NULL


# Down
PA.Down <- Reduce(function(x, y) merge(x, y, all = T), list(
  aECDown, 
  ChPlexDown, 
  HomeoMGDown, 
  MGDown, 
  FibroDown,
  CapECDown)) # merge all lists and keep all unique values
```

```{r}
# Create heatmap

# Up
melt_PA.Up <- melt(PA.Up, id.vars = "CellType", variable.name = "Term", value.name = "FDR") # create long data

Up.OrderedCellTypes <- c( # create specific cell type order that will be used to order the cell types on the X axis
  "Capillary_EC",
  "Arterial_EC",
  "Venous_EC",
  "Microglia",
  "Homeostatic_Microglia",
  "Arterial_SMC",
  "Choroid_Plexus"
)

# create alphabetized term order so all terms from the same source are together on the Y axis. Important if terms are drawn from multiple sources
up.term.order <- t(PA.Up) # transpose 
up.term.order <- as.data.frame(up.term.order) # convert to data frame
up.term.order <- rownames_to_column(up.term.order, "X") # convert rownames to 1st column with "X" as the column title
up.term.order <- subset(up.term.order, select = X) # keep only column X 
up.term.order <- up.term.order[-1,] # remove the 1st column
up.term.order <- str_sort(up.term.order) # sort the object 

# Remove identifiers from melt_PA.Up & up.term.order
melt_PA.Up[c("1", "2")] <- str_split_fixed(melt_PA.Up$Term, "_", 2) # split Term column at the "_"
melt_PA.Up$Term <- NULL # remove the "Term" column
melt_PA.Up$"1" <- NULL # remove the "1" column
melt_PA.Up <- rename(melt_PA.Up,
                     Term = "2") # rename the "2" column back to "Term" - this will no longer have the term IDs attached

up.term.order <- as.data.frame(up.term.order) # convert to data frame
up.term.order[c("1", "2")] <- str_split_fixed(up.term.order$up.term.order, "_", 2) # split up.term.order column at the "_"
up.term.order$up.term.order <- NULL # remove the "up.term.order" column
up.term.order$"1" <- NULL # remove the "1" column
up.term.order <- rename(up.term.order,
                        x = "2")  # rename the "2" column back to "x" - this will no longer have the term IDs attached
up.term.order <- as.matrix(up.term.order) # convert the object to a matrix
up.term.order <- as.character(up.term.order) # conver the object to a character

# Create the upregulated pathways plot using your ordered cell types and alphabetized terms
Up.ggplothmap <- ggplot(melt_PA.Up, aes(x = fct_relevel(CellType, Up.OrderedCellTypes), y = fct_relevel(Term, up.term.order), fill = FDR)) + 
  geom_tile(color = "black") +
  scale_fill_gradientn(colors = topo.colors(5),na.value = "white",limits = c(0, 0.05), breaks = c(0.01, 0.02, 0.03, 0.04, 0.05), label = c(0.01, 0.02, 0.03, 0.04, 0.05)) +
  xlab("") + 
  ylab("") +
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 0.95, vjust = 0.2), axis.text.y = element_text(size = 8, hjust = 1), axis.ticks = element_blank(), panel.background = element_blank()) + 
  scale_x_discrete(expand=c(0,0))
Up.ggplothmap


# Down
melt_PA.Down <- melt(PA.Down, id.vars = "CellType", variable.name = "Term", value.name = "FDR") # create long data

Down.OrderedCellTypes <- c( # create specific cell type order
  "Capillary_EC",
  "Arterial_EC",
  "Microglia",
  "Homeostatic_Microglia",
  "Choroid_Plexus",
  "Fibroblast"
)

down.term.order <- t(PA.Down) # create alphabetized term order
down.term.order <- as.data.frame(down.term.order)
down.term.order <- rownames_to_column(down.term.order, "X")
down.term.order <- subset(down.term.order, select = X)
down.term.order <- down.term.order[-1,]
down.term.order <- str_sort(down.term.order)

# Remove identifiers from melt_PA.Up & up.term.order
melt_PA.Down[c("1", "2")] <- str_split_fixed(melt_PA.Down$Term, "_", 2)
melt_PA.Down$Term <- NULL
melt_PA.Down$"1" <- NULL
melt_PA.Down <- rename(melt_PA.Down,
                     Term = "2")

down.term.order <- as.data.frame(down.term.order)
down.term.order[c("1", "2")] <- str_split_fixed(down.term.order$down.term.order, "_", 2)
down.term.order$down.term.order <- NULL
down.term.order$"1" <- NULL
down.term.order <- rename(down.term.order,
                        x = "2")
down.term.order <- as.matrix(down.term.order)
down.term.order <- as.character(down.term.order)

# Create the downregulated pathways plot using your ordered cell types and alphabetized terms
Down.ggplothmap <- ggplot(melt_PA.Down, aes(x = fct_relevel(CellType, Down.OrderedCellTypes), y = fct_relevel(Term, down.term.order), fill = FDR)) + 
  geom_tile(color = "black") +
  scale_fill_gradientn(colors = topo.colors(5),na.value = "white", limits = c(0, 0.05), breaks = c(0.01, 0.02, 0.03, 0.04, 0.05), label = c(0.01, 0.02, 0.03, 0.04, 0.05)) +
  xlab("") + 
  ylab("") +
  theme(axis.text.x = element_text(size = 8, angle = 90, hjust = 0.95, vjust = 0.2), axis.text.y = element_text(size = 8, hjust = 1), axis.ticks = element_blank(), panel.background = element_blank()) + 
  scale_x_discrete(expand=c(0,0))
Down.ggplothmap
```

# Code below not used...

<!-- ### Load Reactome Data -->
<!-- ```{r} -->
<!-- # Load Analysis from Reactome -->
<!-- Load_File_Path <- -->
<!--   "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/DE_Analysis/Reactome/Condensed/Choroid_Plexus_CSDvsHC_Downregulated.csv" -->
<!-- ReactomeData <- read.csv(file = Load_File_Path) -->

<!-- # set later plot attributes and save names to correspond with cell type UP/DOWN -->
<!-- # title for pathway enrichment plots -->
<!-- title <- "Choroid Plexus: Downregulated" -->

<!-- # save names for plots -->
<!-- GenePlotName <- "Choroid_Plexus_DOWN_PEA_Gene_Ratio.jpeg" -->
<!-- ReactionPlotName <- "Choroid_Plexus_DOWN_PEA_Reaction_Ratio.jpeg" -->
<!-- ``` -->

<!-- ### Clean Up Data -->
<!-- ```{r} -->
<!-- # Rename columns -->
<!-- ReactomeData <- rename(ReactomeData, -->
<!--     Pathway_ID = Pathway.identifier, # new name = old name, -->
<!--     Pathway_Name = Pathway.name, -->
<!--     Genes = X.Entities.found, -->
<!--     Genes_In_Pathway = X.Entities.total, -->
<!--     Genes_Ratio = Entities.ratio, -->
<!--     p_val = Entities.pValue, -->
<!--     FDR = Entities.FDR, -->
<!--     Reactions = X.Reactions.found, -->
<!--     Reactions_In_Pathway = X.Reactions.total, -->
<!--     Reactions_Ratio = Reactions.ratio, -->
<!--     Species_ID = Species.identifier, -->
<!--     Species_Name = Species.name, -->
<!--     Genes_Found_Names = Submitted.entities.found, -->
<!--     Mapped_Entities = Mapped.entities, -->
<!--     Found_Reaction_IDs = Found.reaction.identifiers -->
<!--   ) -->

<!-- # Keep only pathways with FRD < 0.05 -->
<!-- Sig.ReactomeData <- subset(ReactomeData, FDR < 0.05) -->
<!-- ``` -->

<!-- ### Plot enriched pathways -->
<!-- ```{r} -->
<!-- # Focus on Gene Ratio -->
<!-- geneRatio <- ggplot(Sig.ReactomeData, -->
<!--             aes(x = Genes_Ratio, y = reorder(Pathway_Name, Genes_Ratio), colour = FDR, size = Genes)) + -->
<!--   geom_point() + -->
<!--   scale_colour_gradientn(colours = rainbow(5), guide = "colourbar") + -->
<!--   expand_limits(color = seq(0, 0.05, by = 0.05)) + -->
<!--   ggtitle(title) + xlab("Gene Ratio") + ylab("Pathway") + -->
<!--   xlim(0, NA) + -->
<!--   theme_bw() + theme(axis.text = element_text(color = "black", size = 10)) -->
<!-- geneRatio -->


<!-- # Focus on Reactions Ratio -->
<!-- reactionRatio <- ggplot(Sig.ReactomeData, -->
<!--             aes(x = Reactions_Ratio, y = reorder(Pathway_Name, Reactions_Ratio), colour = FDR, size = Reactions)) + -->
<!--   geom_point() + -->
<!--   scale_colour_gradientn(colours = rainbow(5), guide = "colourbar") + -->
<!--   expand_limits(color = seq(0, 0.05, by = 0.05)) + -->
<!--   ggtitle(title) + xlab("Reactions Ratio") + ylab("Pathway") + -->
<!--   xlim(0, NA) + -->
<!--   theme_bw() + theme(axis.text = element_text(color = "black", size = 10)) -->
<!-- reactionRatio -->
<!-- ``` -->

<!-- ### Save Plots -->
<!-- ```{r} -->
<!-- Save_File_Path <- -->
<!--   "D:/Research/NIH/Herkenham_Lab/10X_Neurovascular_scRNAseq/Plots/Reactome/Mouse_DE_Analysis/Pathway_Enrichment" -->

<!-- # save geneRatio plot -->
<!-- ggsave(filename = GenePlotName, plot = geneRatio, device = "jpeg", path = Save_File_Path, units = "in", width = 12, height = 10) -->

<!-- # save reactionRatio Plot -->
<!-- ggsave(filename = ReactionPlotName, plot = reactionRatio, device = "jpeg", path = Save_File_Path, units = "in", width = 12, height = 10) -->

<!-- ``` -->
